---
title: "PM 566 Final Project - California Supplier Diversity and Net Income"
author: "Norma Marshall"
format: html
embed-resources: true
warning: false
page-layout: full
---

```{r, echo = FALSE}
library(dplyr)
library(readxl)
library(tidygeocoder)
library(leaflet)
library(tidyr)
library(knitr)
library(ggplot2)
library(htmltools)
library(tidytext)
library(readr)
library(stringr)    
library(forcats)  
library (plotly)
library(writexl)
library(shiny)
library(viridis)

```

# Background and Research Question

To accelerate efforts to reduce health care disparities, hospitals and health systems increase their efforts in core areas such as staff and leadership diversity and cultural competence. However, the economic relationship between healthcare systems and the communities they serve are growing in importance.

Supplier diversity refers to when an organization procures goods and services from a variety of businesses, including those that are at least 51% owned, managed, and operated from marginalized and minority groups. These include women, veterans, African Americans, LGBTQIA+, and more. The private sector, including the healthcare industry has incorporated supplier diversity programs into their business practices after historically being adopted by the federal government and its contractors. According to the Harvard Business Review, supplier diversity programs are important in combatting social injustice and systemic racism in the US as they actively include diverse representation and inclusion in hospital operations and supply chains. In addition, for the moral and ethical arguments, supplier diversity programs have commercial in hospitals and health systems. These benefits include greater innovation and value through cost reductions, expansions of external partnerships, local job creation, better understanding of supply chain sourcing process and sources, and easier compliance with government and grant contracts.

On an annual basis (individual hospital fiscal year), individual hospitals and systems report detailed facility level financial data to the Department of Health Care Access and Information (HCAI). This data includes detailed facility level data on services capacity, inpatient/outpatient utilization, revenues, and expenses by type and payer. In addition, the Health and Safety Code Section 1339.85-1339.87 requires individual hospitals with operating expenses over $50 million to report hospital supplier and diversity reports explaining the hospitals’ supplier diversity statement and procurement efforts regarding minority, women, LGBT, and disabled veteran enterprises. Each hospital also reports tier 1 and tier 2 spend for each category of diverse suppliers. Tier 1 spend refers to the total amount of procurement dollars that an organization spends directly with a certified diverse supplier. Tier 2, on the other hand represents the total amount of procurement dollars that an organization spends with a given supplier, who then sources products and services from another certified diverse supplier. Tier 2 relationships enable hospitals to increase diversity spend while working with existing suppliers or with suppliers who may possess rare capabilities.


Medi-Cal eligible patients are also a significant portion of the population in California, making up a large portion of the state’s healthcare system. The HCAI’s financial report also reveals the revenue generated by hospitals from Medi-Cal patients, offering valuable insights into the economic pressures that hospitals face while serving this demographic. As the hospitals system increasingly rely on reimbursements from Medi-Cal, it is important to examine how hospitals' procurement practices might influence their financial performance, including revenue from Medi-Cal patients.

This report merges the annual financial data and supplier diversity reports for 2023 to answer the question **Are California hospitals with diverse suppliers profitable?** Supplier diversity aims to increase innovation and drive down prices for supplies and goods though competition while also aiming to improve health equity and combat social injustice in the US through business practices. This exploratory data analysis aims to see if funds dedicated to goods and services from diversely owned businesses can lead to better financial outcomes.

# Methods

A novel dataset was collected by merging two data sets, HCAI's Hospital Annual Financial Disclosure Report for 2023, and HCAI’s Supplier Diversity report from 2023. Required by state law, supplier diversity and financial data are reported each year to the HCAI. Datasets were merged on the shared hospital name variable to create a comprehensive data containing both supplier diversity and financial metrics for each hospital.

Following the merge, several variables indicating procurement from diverse backgrounds were recoded to be used as numeric variables, allowing for quantitative analysis. Address related variables were recoded into latitude and longitude variables to facilitate geocoding using the tidygeocoder package for recoding and the leaflet package for visualization.

Frequency tables were generated to view top hospitals in net income, supplier procurement from different backgrounds, and demographic data. Maps were generated to locate the top performing hospitals and correlations were run to quantify the relationship between supplier procurement and hospital net income.Pearson’s correlation coefficient was then calculated to assess the linear relationship between procurement ratios for each diversity category and net and Medi-Cal revenue. This quantified the economic impact of diversity procurement on hospital revenue. The correlation analysis allowed for the identification of any significant relationships between hospitals that allocate a higher percentage of procurement to minority suppliers and their corresponding medical revenue..




```{r, echo = FALSE}


# Read Excel file



# Safely converting columns to numeric, handling non-numeric values
final_hospitals <- read_excel("final_hospitals.xlsx") %>%
  mutate(across(
    c(Tier_I_African_American, Tier_II_African_American, Total_African_American,
      Tier_I_Hispanic_American, Tier_II_Hispanic_American, Total_Hispanic_American,
      Tier_I_Native_American, Tier_II_Native_American, Total_Native_American,
      Tier_I_Asian_Pacific_American, Tier_II_Asian_Pacific_American, Total_Asian_Pacific_American,
      Tier_I_Unknown_Minority, Tier_II_Unknown_Minority, Total_Unknown_Minority,
      Total_Tier_I_Minority, Total_Tier_II_Minority, Total_Minority,
      Tier_I_Women, Tier_II_Women, Total_Women,
      Tier_I_LGBT, Tier_II_LGBT, Total_LGBT,
      Tier_I_Disabled_Veteran, Tier_II_Disabled_Veteran, Total_Disabled_Veteran,
      Tier_I_Less_Duplicated_Amount, Tier_II_Less_Duplicated_Amount, Total_Less_Duplicated_Amount,
      Combined_Tier_I_Total, Combined_Tier_II_Total, Combined_Total, Total_Hospital_Procurement, NETRV_MCAL_TR,
      latitude...73, longitude...74),  # Adding latitude and longitude to the list of columns to convert
    ~as.numeric(gsub(",", "", .)), .names = "numeric_{.col}")
  )

 final_hospitals <- final_hospitals %>%
  rename(
    latitude = `latitude...73`,
    longitude = `longitude...74`
  )
```

# Demographics

In this report there are 372 hospitals with both financial and supplier diversity data reported the to the HCAI in 2023. On average these hospitals report earning over 24 million dollars over the course of the year and spend over \$9.6 million on supplies from diverse suppliers, totaling about 9 percent of the total dollars spent on the procurement on supplies annually.

A majority of these hospitals (320) serve urban areas in California, as demonstrated by large clusters of hospitals around major cities such as San Francisco, San Diego, and Los Angeles. Over 53% of the hospitals are nonprofit, including church related facilities. Average dollars spent on Tier 1, Tier 2, and total suppliers by each minority group were also reported (Tables 1–4).


```{r, echo = FALSE}



summary_table <- final_hospitals %>%
  summarize(
    `Total Hospitals` = n(),
    `Average Procurement from Diverse Suppliers` = mean(Combined_Total, na.rm = TRUE),
    `Average Total Hospital Procurement` = mean(Total_Hospital_Procurement, na.rm = TRUE),
    `Average Net Income` = mean(NET_INCOME, na.rm = TRUE)
  )

kable(summary_table, caption = "Table 1. Summary Table of Hospital Data")


urban_rural_freq <- final_hospitals %>%
  group_by(MSSA) %>%
  summarize(Frequency = n()) %>%
  ungroup()

type_control_freq <- final_hospitals %>%
  group_by(Type_Control) %>%
  summarize(Frequency = n()) %>%
  ungroup()

kable(urban_rural_freq, caption = "Table 2. Frequencies of Urban/Rural Hospitals")
kable(type_control_freq, caption = "Table 3. Frequencies of Hospital Types)")

```

```{r, echo = FALSE}
# Define Tier I and Tier II variables
tier1_vars <- c(
  "Tier_I_African_American",
  "Tier_I_Hispanic_American",
  "Tier_I_Native_American",
  "Tier_I_Asian_Pacific_American",
  "Tier_I_Unknown_Minority",
  "Total_Tier_I_Minority",
  "Tier_I_Women",
  "Tier_I_LGBT",
  "Tier_I_Disabled_Veteran",
  "Tier_I_Less_Duplicated_Amount",
  "Combined_Tier_I_Total"
)

tier2_vars <- c(
  "Tier_II_African_American",
  "Tier_II_Hispanic_American",
  "Tier_II_Native_American",
  "Tier_II_Asian_Pacific_American",
  "Tier_II_Unknown_Minority",
  "Total_Tier_II_Minority",
  "Tier_II_Women",
  "Tier_II_LGBT",
  "Tier_II_Disabled_Veteran",
  "Tier_II_Less_Duplicated_Amount",
  "Combined_Tier_II_Total"
)

# Calculate average procurement for each group in Tier 1
avg_tier1_groups <- sapply(tier1_vars, function(var) {
  mean(final_hospitals[[var]], na.rm = TRUE)
})

# Calculate average procurement for each group in Tier 2
avg_tier2_groups <- sapply(tier2_vars, function(var) {
  mean(final_hospitals[[var]], na.rm = TRUE)
})

# Combine results into a data frame for better readability
avg_procurement <- data.frame(
  Group = c(tier1_vars, tier2_vars),
  Average_Procurement = c(avg_tier1_groups, avg_tier2_groups),
  Tier = rep(c("Tier 1", "Tier 2"), c(length(tier1_vars), length(tier2_vars)))
)



# Assuming your current table is named `avg_procurement`

# Extract ethnicities by removing prefixes and other identifiers
avg_procurement$Ethnicity <- gsub(
  "Tier_I_|Tier_II_|Combined_|Total_|Less_Duplicated_Amount", 
  "", 
  avg_procurement$Group
)

# Pivot the data to have separate columns for Tier 1 and Tier 2 averages
library(dplyr)
library(tidyr)

final_table <- avg_procurement %>%
  select(Ethnicity, Average_Procurement, Tier) %>%
  pivot_wider(
    names_from = Tier,
    values_from = Average_Procurement,
    names_prefix = "Tier_"
  ) %>%
  # Calculate Total Average
  mutate(Total_Average = rowMeans(select(., starts_with("Tier_")), na.rm = TRUE))

# Rename the columns based on the actual column names
final_table <- final_table %>%
  rename(
    `Tier 1 Average` = `Tier_Tier 1`,
    `Tier 2 Average` = `Tier_Tier 2`
  )

# Arrange and display the table
kable(final_table, caption = "Table 4. Average Procrument Per Diverse Supplier (Tier 1/2/Total) ")


```



# How much are hosptials spending on supplies from minority owned businessess?

Washington Hospital in Fremont spends in the most on minority owned suppliers, followed by Stanford Health Care. These two hospitals hwoever, are major outliers compared to other hospitals in California. The interactive figure below omits these two hospitals to better depict the distribution of hospitals spending and spending n minority owned businesses. However its important to realize that any possible correlations could be driven by these two major hospitals.  In specific categories, Kaiser Permanente in Santa Clara spends the most on African American suppliers while Stanford leads for Hispanic and Asian/Pacific Category. Washington Hospital in Fremont also is the top performer in the unknown minority category while, such as Kaiser Foundation Hospital - San Diego and Contra Costa Regional Medical Center, lead in categories like total women-owned and LGBT-owned suppliers, respectively (Table 5). 

Table 6 shows the top-performing hospitals in terms of procurement from diverse suppliers, with Washington Hospital - Fremont leading in both "Total Unknown Minority" (76.71%) and "Total Minority" (80.61%). Langley Porter Psychiatric Institute excels in procurement from Asian Pacific American suppliers, accounting for 64.61% of its total procurement. Other hospitals, such as Providence Little Company of Mary MC - San Pedro and Kaiser Foundation Hospital - Modesto, show strong commitment to supporting women and disabled veteran-owned businesses, with 76.16% and 28.84% of procurement from these groups, respectively.
```{r, echo = FALSE}
# Filter out the specified hospitals
filtered_data <- final_hospitals %>%
  filter(!Hospital_Name %in% c("STANFORD HEALTH CARE", "WASHINGTON HOSPITAL - FREMONT"))

base_plot <- ggplot(filtered_data, aes(x = Combined_Total, y = NET_INCOME, text = paste("Hospital Name: ", Hospital_Name, "<br>Net Income: $", NET_INCOME, "Diverse Procrument: $", Combined_Total))) +
  geom_point(color = "blue", size = 2, alpha = 0.7) +
  labs(
    title = "Interactive Scatter Plot of Combined Total vs. Net Income",
    x = "Total Procurement from Diverse Suppliers",
    y = "Net Income"
  ) +
  theme_minimal()

# Convert the ggplot object to a plotly interactive plot
interactive_plot <- ggplotly(base_plot, tooltip = "text")

# Display the plot
interactive_plot


```

```{r, echo = FALSE}

# Get the top diverse hospitals and select only the relevant columns
top_diverse_hospitals <- final_hospitals %>%
  arrange(desc(Total_Minority)) %>%  
  select(Hospital_Name, Total_Minority) %>%  
  head(n = 10)  

kable(top_diverse_hospitals, col.names = c("Hospital Name", "Combined Total"), caption = "Table 5. Combined Total Spent on Minority Owned Suppliers" )


```

```{r, echo = FALSE}
selected_vars <- c(
  "Total_African_American", "Total_Hispanic_American",
  "Total_Native_American", "Total_Asian_Pacific_American",
  "Total_Unknown_Minority", "Total_Minority",
  "Total_Women", "Total_LGBT", "Total_Disabled_Veteran"
)

top_hospitals_df <- data.frame(Top_Hospital = character(), Top_Value = numeric(), Category = character(), stringsAsFactors = FALSE)

# Loop through each variable to find the top hospital based on the percentage
for (var_name in selected_vars) {
  top_hospital <- final_hospitals %>%
    filter(!is.na(!!sym(var_name))) %>%  # Exclude NA values for the variable
    mutate(Percentage = (!!sym(var_name) / Total_Hospital_Procurement) * 100) %>%  # Calculate the percentage
    filter(Percentage <= 99.999) %>%  # Exclude rows where the percentage is over 100
    top_n(1, Percentage) %>%  # Get the hospital with the highest percentage
    select(Hospital_Name, Percentage) %>%  # Select relevant columns
    rename(Top_Hospital = Hospital_Name, Top_Value = Percentage) %>%  # Rename for clarity
    mutate(Category = var_name)  # Add the category as a new column
  
  top_hospitals_df <- rbind(top_hospitals_df, top_hospital)
}

# Display the table with kable, including the Category column
kable(top_hospitals_df, 
      col.names = c("Top Hospital", "% of Total Procurement", "Category"),
      caption = "Table 6. Top Performing Hospitals by Category (percentage of of total procurment from diverse suppliers)")


```

# Where are the most diverse hopstials located? 

Across categories and in total, a majority of hospitals that spend the most money on diverse suppliers reside in the bay  or Los Angeles  areas However, there are a significant number of hospitals that spend a lot of money suppliers from women and disabled veterans in Southern California. Hospitals in northern California include a few state and University of California hospitals including UC San Francisco and Stanford (Refer to Interactive figure below). 




::: {.panel-tabset}

## Total Minority Groups 

```{r, echo=FALSE}

total_minority_palette <- colorNumeric(palette = viridis(256), 
                                       domain = final_hospitals$Total_Minority)

# Create a leaflet plot for hospitals
leafplot <- leaflet(final_hospitals) %>% 
  addProviderTiles('CartoDB.Positron') %>% 
  addCircles(
    lat = ~latitude,  
    lng = ~longitude,  
    label = ~paste0("Hospital: ", Hospital_Name, "<br>Total Minority: ", Total_Minority),  
    color = ~total_minority_palette(Total_Minority),  
    opacity = 1, 
    fillOpacity = 1,
    stroke = FALSE,  
    radius = 5       
  ) %>%
  addLegend('bottomleft', 
            pal = total_minority_palette, 
            values = final_hospitals$Total_Minority,
            title = 'Total Minority in Hospitals', 
            opacity = 1) %>%
  setView(lng = -119.4179, lat = 36.7783, zoom = 6)  # Set view to California

leafplot


```

## Total African American 
```{r, echo=FALSE}

# Load the viridis package
library(viridis)

# Create a viridis color palette for Total_African_American
african_american_palette <- colorNumeric(palette = viridis(256), 
                                         domain = final_hospitals$Total_African_American)

# Create a leaflet plot for hospitals
african_american_map <- leaflet(final_hospitals) %>% 
  addProviderTiles('CartoDB.Positron') %>% 
  addCircles(
    lat = ~latitude,  
    lng = ~longitude,  
    label = ~paste0("Hospital: ", Hospital_Name, "<br>Total African American: ", Total_African_American),  
    color = ~african_american_palette(Total_African_American),  
    opacity = 1, 
    fillOpacity = 1,
    stroke = FALSE,  
    radius = 5       
  ) %>%
  addLegend('bottomleft', 
            pal = african_american_palette, 
            values = final_hospitals$Total_African_American,
            title = 'Total African American in Hospitals', 
            opacity = 1) %>%
  setView(lng = -119.4179, lat = 36.7783, zoom = 6)  # Set view to California

african_american_map

```


## Hispanic American 
```{r, echo=FALSE}
# Create a viridis color palette for Total_Hispanic_American
hispanic_american_palette <- colorNumeric(palette = viridis(256), 
                                          domain = final_hospitals$Total_Hispanic_American)

# Create a leaflet plot for hospitals
hispanic_american_map <- leaflet(final_hospitals) %>% 
  addProviderTiles('CartoDB.Positron') %>% 
  addCircles(
    lat = ~latitude,  
    lng = ~longitude,  
    label = ~paste0("Hospital: ", Hospital_Name, "<br>Total Hispanic American: ", Total_Hispanic_American),  
    color = ~hispanic_american_palette(Total_Hispanic_American),  
    opacity = 1, 
    fillOpacity = 1,
    stroke = FALSE,  
    radius = 5       
  ) %>%
  addLegend('bottomleft', 
            pal = hispanic_american_palette, 
            values = final_hospitals$Total_Hispanic_American,
            title = 'Total Hispanic American in Hospitals', 
            opacity = 1) %>%
  setView(lng = -119.4179, lat = 36.7783, zoom = 6)  # Set view to California

hispanic_american_map
```

## Total Native American 
```{r, echo=FALSE}
# Create a viridis color palette for Total_Hispanic_American
hispanic_american_palette <- colorNumeric(palette = viridis(256), 
                                          domain = final_hospitals$Total_Hispanic_American)

# Create a leaflet plot for hospitals
hispanic_american_map <- leaflet(final_hospitals) %>% 
  addProviderTiles('CartoDB.Positron') %>% 
  addCircles(
    lat = ~latitude,  
    lng = ~longitude,  
    label = ~paste0("Hospital: ", Hospital_Name, "<br>Total Hispanic American: ", Total_Hispanic_American),  
    color = ~hispanic_american_palette(Total_Hispanic_American),  
    opacity = 1, 
    fillOpacity = 1,
    stroke = FALSE,  
    radius = 5       
  ) %>%
  addLegend('bottomleft', 
            pal = hispanic_american_palette, 
            values = final_hospitals$Total_Hispanic_American,
            title = 'Total Hispanic American in Hospitals', 
            opacity = 1) %>%
  setView(lng = -119.4179, lat = 36.7783, zoom = 6)  # Set view to California

hispanic_american_map

```

## Native American 

```{r, echo=FALSE}
# Create a viridis color palette for Total_Hispanic_American
hispanic_american_palette <- colorNumeric(palette = viridis(256), 
                                          domain = final_hospitals$Total_Hispanic_American)

# Create a leaflet plot for hospitals
hispanic_american_map <- leaflet(final_hospitals) %>% 
  addProviderTiles('CartoDB.Positron') %>% 
  addCircles(
    lat = ~latitude,  
    lng = ~longitude,  
    label = ~paste0("Hospital: ", Hospital_Name, "<br>Total Hispanic American: ", Total_Hispanic_American),  
    color = ~hispanic_american_palette(Total_Hispanic_American),  
    opacity = 1, 
    fillOpacity = 1,
    stroke = FALSE,  
    radius = 5       
  ) %>%
  addLegend('bottomleft', 
            pal = hispanic_american_palette, 
            values = final_hospitals$Total_Hispanic_American,
            title = 'Total Hispanic American in Hospitals', 
            opacity = 1) %>%
  setView(lng = -119.4179, lat = 36.7783, zoom = 6)  # Set view to California

hispanic_american_map

```

## Asian Pacific American 

```{r, echo=FALSE}
# Create a viridis color palette for Total_Asian_Pacific_American
asian_pacific_american_palette <- colorNumeric(palette = viridis(256), 
                                               domain = final_hospitals$Total_Asian_Pacific_American)

# Create a leaflet plot for hospitals
asian_pacific_american_map <- leaflet(final_hospitals) %>% 
  addProviderTiles('CartoDB.Positron') %>% 
  addCircles(
    lat = ~latitude,  
    lng = ~longitude,  
    label = ~paste0("Hospital: ", Hospital_Name, "<br>Total Asian Pacific American: ", Total_Asian_Pacific_American),  
    color = ~asian_pacific_american_palette(Total_Asian_Pacific_American),  
    opacity = 1, 
    fillOpacity = 1,
    stroke = FALSE,  
    radius = 5       
  ) %>%
  addLegend('bottomleft', 
            pal = asian_pacific_american_palette, 
            values = final_hospitals$Total_Asian_Pacific_American,
            title = 'Total Asian Pacific American in Hospitals', 
            opacity = 1) %>%
  setView(lng = -119.4179, lat = 36.7783, zoom = 6)  # Set view to California

asian_pacific_american_map
```

## Unknown Minority

```{r, echo=FALSE}
# Create a viridis color palette for Total_Unknown_Minority
unknown_minority_palette <- colorNumeric(palette = viridis(256), 
                                         domain = final_hospitals$Total_Unknown_Minority)

# Create a leaflet plot for hospitals
unknown_minority_map <- leaflet(final_hospitals) %>% 
  addProviderTiles('CartoDB.Positron') %>% 
  addCircles(
    lat = ~latitude,  
    lng = ~longitude,  
    label = ~paste0("Hospital: ", Hospital_Name, "<br>Total Unknown Minority: ", Total_Unknown_Minority),  
    color = ~unknown_minority_palette(Total_Unknown_Minority),  
    opacity = 1, 
    fillOpacity = 1,
    stroke = FALSE,  
    radius = 5       
  ) %>%
  addLegend('bottomleft', 
            pal = unknown_minority_palette, 
            values = final_hospitals$Total_Unknown_Minority,
            title = 'Total Unknown Minority in Hospitals', 
            opacity = 1) %>%
  setView(lng = -119.4179, lat = 36.7783, zoom = 6)  # Set view to California

unknown_minority_map

```
:::





# What hosptials make the most money?

Regardless of supplier diversity, the top earning hosptials are in the bay area or Los Angeles/Orange County areas. Cedars_Sinai, LA General, UCLA, and Childrens Hopsital are in the top 10 earning hospitals in California, however this does not mean they spend the most on supplies from minority owned businesses. (Table 7)

Most hospitals in the plot below net the average amount in income every year, with some like Eden Medical Center earning the least amount of income.

```{r, echo = FALSE}

# Load the knitr package if it’s not already loaded
library(knitr)

# Prepare the top hospitals table
top_hospitals <- final_hospitals %>%
  arrange(desc(NET_INCOME)) %>%  
  head(n = 10) %>%  
  select(Hospital_Name, NET_INCOME)

# Display the table with just kable
kable(top_hospitals, 
      col.names = c("Hospital Name", "Net Income"), 
      caption = "Table 7. Top 10 Hospitals by Net Income")


```

# Is diverse supplier procurement correlated with net hospital income?

To investigate whether or not the relationship between supplier diversity and net income for a hospital can vary between these two types of procurement from diverse suppliers. Correlations were ran between net income and supplier procurement.

There is moderate positive correlation between Tier 1 suppliers that are Hispanic, and women owned (0.59,0.60), as well a s total procurement from Hispanic owned suppliers. There is also a slight negative correlation between tier two supplier procurement from suppliers that are women owned, LGBT owned, and overall (\<0) suggesting a negative correlation between increasing tier 2 procurement form diverse suppliers and net income. (Table 8)

```{r, echo = FALSE}
selected_vars <- c(
  "Total_African_American", "Total_Hispanic_American",
  "Total_Native_American", "Total_Asian_Pacific_American",
  "Total_Unknown_Minority", "Total_Minority",
  "Total_Women", "Total_LGBT", "Total_Disabled_Veteran",
  "NET_INCOME"
)

# Define Tier I and Tier II variables
tier1_vars <- c(
  "Tier_I_African_American",
  "Tier_I_Hispanic_American",
  "Tier_I_Native_American",
  "Tier_I_Asian_Pacific_American",
  "Tier_I_Unknown_Minority",
  "Total_Tier_I_Minority",
  "Tier_I_Women",
  "Tier_I_LGBT",
  "Tier_I_Disabled_Veteran",
  "Tier_I_Less_Duplicated_Amount",
  "Combined_Tier_I_Total"
)

tier2_vars <- c(
  "Tier_II_African_American",
  "Tier_II_Hispanic_American",
  "Tier_II_Native_American",
  "Tier_II_Asian_Pacific_American",
  "Tier_II_Unknown_Minority",
  "Total_Tier_II_Minority",
  "Tier_II_Women",
  "Tier_II_LGBT",
  "Tier_II_Disabled_Veteran",
  "Tier_II_Less_Duplicated_Amount",
  "Combined_Tier_II_Total"
)

# Combine Tier I and Tier II variables into one list
all_vars <- c(tier1_vars, tier2_vars, selected_vars)

# Function to calculate correlations for a specific ethnic group
calculate_group_correlations <- function(group_vars, net_income_col) {
  # Create a correlation matrix for the ethnic group and NET_INCOME
  correlation_data <- final_hospitals %>%
    select(all_of(c(group_vars, net_income_col))) %>%
    filter(complete.cases(.))  # Remove rows with any NA values
  
  # Calculate the correlation matrix
  correlation_matrix <- cor(correlation_data, use = "complete.obs")
  
  # Extract correlations with NET_INCOME
  cor_with_net_income <- correlation_matrix[net_income_col, group_vars]
  
  return(cor_with_net_income)
}

# Calculate correlations for each ethnic group
ethnicity_groups <- c(
  "African_American", "Hispanic_American", "Native_American", 
  "Asian_Pacific_American", "Unknown_Minority", "Minority", 
  "Women", "LGBT", "Disabled_Veteran"
)

# Create empty vectors to store the correlation values
tier1_correlations <- numeric(length(ethnicity_groups))
tier2_correlations <- numeric(length(ethnicity_groups))
overall_correlations <- numeric(length(ethnicity_groups))

# Loop through each ethnicity and calculate correlations
for (i in 1:length(ethnicity_groups)) {
  # Get the specific ethnic group variables for Tier 1 and Tier 2
  tier1_ethnicity_vars <- grep(ethnicity_groups[i], tier1_vars, value = TRUE)
  tier2_ethnicity_vars <- grep(ethnicity_groups[i], tier2_vars, value = TRUE)
  
  # Calculate correlations for Tier 1, Tier 2, and Total ethnic group (overall)
  tier1_correlations[i] <- calculate_group_correlations(tier1_ethnicity_vars, "NET_INCOME")
  tier2_correlations[i] <- calculate_group_correlations(tier2_ethnicity_vars, "NET_INCOME")
  
  # Calculate overall correlation for the ethnic group
  overall_ethnicity_vars <- c(tier1_ethnicity_vars, tier2_ethnicity_vars)
  overall_correlations[i] <- calculate_group_correlations(overall_ethnicity_vars, "NET_INCOME")
}

# Create a summary table
correlation_table <- data.frame(
  Ethnicity = ethnicity_groups,
  Tier_1_Correlation = tier1_correlations,
  Tier_2_Correlation = tier2_correlations,
  Total_Average_Correlation = overall_correlations
)

# Display the table using kable
kable(correlation_table, 
      col.names = c("Ethnicity", "Tier 1 Correlation", "Tier 2 Correlation", "Total Average Correlation"),
      caption = "Table 8. Correlation of Diverse Groups with Net Income in Tier 1, Tier 2, and Overall Procurement"
)


```

## What about Medi-Cal Revenue?

Medi-Cal revenue is an important factor to consider when looking at procurement from diverse suppliers because hospitals that serve a higher proportion of Medi-Cal patients may face different financial pressures compared to those with a more commercially insured patient base. The correlation analysis shows that the relationships between diverse supplier procurement and Medi-Cal revenue are mostly weak and negative for several minority groups, indicating limited impact or a lack of strong association. However, procurement from diverse suppliers can still be a strategy to fulfill diversity and inclusion goals, potentially improve cost efficiency, foster local engagement, and ensure compliance with government contracts, which may influence financial outcomes and help advance community health equity objectives. (Table 9)

```{r, echo = FALSE}
# Function to calculate correlation with medical revenue for each selected variable ratio
calculate_ratio_correlation <- function(group_vars, medical_revenue_col, procurement_col) {
  
  # Create an empty vector to store correlation values
  correlation_values <- numeric(length(group_vars))
  
  # Loop through each group variable (e.g., Total_African_American, Total_Hispanic_American)
  for (i in 1:length(group_vars)) {
    selected_var <- group_vars[i]
    
    # Ensure the selected variable exists in the dataset
    if (!(selected_var %in% colnames(final_hospitals))) {
      stop(paste("Column", selected_var, "does not exist in the dataset"))
    }
    
    # Calculate the ratio of the selected variable to Total_Hospital_Procurement
    final_hospitals <- final_hospitals %>%
      mutate(Ratio = .[[selected_var]] / .[[procurement_col]])  # Create the Ratio column
    
    # Select columns for correlation calculation: Ratio and medical revenue
    correlation_data <- final_hospitals %>%
      select(all_of(c(selected_var, "Ratio", medical_revenue_col))) %>%
      filter(complete.cases(.))  # Remove rows with any NA values
    
    # Calculate the correlation between Ratio and MEDICAL_REVENUE
    correlation_matrix <- cor(correlation_data, use = "complete.obs")
    
    # Extract the correlation between the Ratio and MEDICAL_REVENUE
    correlation_values[i] <- correlation_matrix["Ratio", medical_revenue_col]
  }
  
  # Return the correlation values
  return(correlation_values)
}

# Define the selected variables and medical revenue column
selected_vars <- c(
  "Total_African_American", "Total_Hispanic_American", "Total_Native_American", 
  "Total_Asian_Pacific_American", "Total_Unknown_Minority", "Total_Minority", 
  "Total_Women", "Total_LGBT", "Total_Disabled_Veteran"
)

# Define the procurement column
procurement_col <- "Total_Hospital_Procurement"

# Define the medical revenue column
medical_revenue_col <- "NETRV_MCAL_TR"  # Or your specific column name for medical revenue

# Call the function to get correlation values for each selected variable
correlation_results <- calculate_ratio_correlation(selected_vars, medical_revenue_col, procurement_col)

# Create a summary table with the results
correlation_table <- data.frame(
  Variable = selected_vars,
  Correlation_with_Medical_Revenue = correlation_results
)

# Display the table using kable
kable(correlation_table, 
      col.names = c("Variable", "Correlation with Medical Revenue"),
      caption = "Table 9. Correlation of Percentage of total Procurment from Diversie Suppliers with Medi-Cal Net Revenue"
)


```

# What are hospitals saying in their commitment to supplier diversity?

Each hospital is also required to report a supplier diversity statement to state their commitment to procuring from diverse suppliers. W To identify common themes, I highlighted the top 20 unique words and 3-word phrases to identify common themes in supplier diversity statements.

From this analysis suppliers have a strong commitment to procuring from diverse suppliers in order to support the business needs of the hospitals. They also aim to drive some sort of competition, possibly between suppliers to lower prices.

It is also important to note that many individuals submitting these reports on the behalf of hospitals may not choose to submit a diversity statement, but reported ~something~ to meet the requirement,  thus many entries may have  smaller nonsensicle terms like "N/A" or "relies on the University of California procurement policies". 

::: {.panel-tabset}

## Top Terms Found in Supplier Diversity Statements

```{r, echo = FALSE}
final_hospitals %>%
  unnest_tokens(word, Supplier_Diversity_Statement) %>%
   filter(!str_detect(word, "\\d")) %>% 
   anti_join(stop_words, by = c("word"))%>%
  count(word, sort=TRUE) %>%
  top_n(20, n) %>%
  ggplot(aes(x = n, y = fct_reorder(word, n))) +
  geom_col()+
   ggtitle("Top 20 Terms Found In Supplier Diversity Statements")

```

## Top Tokens

```{r, echo=FALSE}

final_hospitals %>%
  unnest_ngrams(token, Supplier_Diversity_Statement, n = 3) %>%
  count(token, sort=TRUE) %>%
  top_n(20, n) %>%
  ggplot(aes(x = n, y = fct_reorder(token, n))) +
  geom_col()+
  ggtitle("Top 20 Tokens Found In Supplier Diversity Statements")
```

## Table 10. Top 5 TF-IDF values by hospital type

```{r, echo = FALSE}
# Load necessary libraries
library(tidyverse)
library(tidytext)
library(knitr)
library(SnowballC)  #
library(stringr)
library(stopwords)

# Clean and preprocess suppleir diversity
final_hospitals_clean <- final_hospitals %>%
  mutate(Supplier_Diversity_Statement = tolower(Supplier_Diversity_Statement)) %>%
  mutate(Supplier_Diversity_Statement = str_remove_all(Supplier_Diversity_Statement, "[^[:alpha:][:space:]]")) %>%
  mutate(Supplier_Diversity_Statement = str_squish(Supplier_Diversity_Statement)) %>%
  mutate(Supplier_Diversity_Statement = sapply(Supplier_Diversity_Statement, function(x) {
    words <- str_split(x, " ")[[1]]
    words_clean <- words[!words %in% stopwords("en")]
    paste(words_clean, collapse = " ")
  }))


# Tokenize the Supplier_Diversity_Statement column grouped by Type_Control
tf_idf_grouped <- final_hospitals_clean %>%
  unnest_tokens(word, Supplier_Diversity_Statement) %>%   # Tokenizing the words from Supplier_Diversity_Statement column
  count(Type_Control, word) %>%                          # Count words grouped by Type_Control
  group_by(Type_Control) %>%                             # Group by Type_Control
  bind_tf_idf(word, Type_Control, n) %>%                  # Calculate TF-IDF within each Type_Control group
  arrange(desc(tf_idf)) %>%                              # Sort by TF-IDF scores
  group_by(Type_Control) %>%
  top_n(5, tf_idf)                                      # Get the top 5 words per each Type_Control group

# Display the result in a table
kable(tf_idf_grouped, caption = " Table 10. Top 5 TF-IDF Values in Supplier Diversity Statements by Hospital Type ")

```

:::

## Supplier Diversity Statments 

If you want to see the supplier diversity statement for a specific hopsital of your choosing you can enter in the search bar below! 

```{r, echo = FALSE}
library(DT)
library(dplyr)

# Assuming final_hospitals is your dataset with columns "Hospital_Name" and "Supplier_Diversity_Statement"

# Create the interactive table
datatable(final_hospitals %>% select(Hospital_Name, Supplier_Diversity_Statement), 
          options = list(pageLength = 5, autoWidth = TRUE), 
          caption = "Table 11. Hospital Supplier Diversity Statements")


```

# Conclusion

Overall, preliminary data suggest a potential positive relationship between procurement from supplier diversity and increased income for hospitals. This remains a critical priority for hospitals and medical centers throughout California. Notably, some of the highest-earning hospitals are not allocating significant resources to diverse suppliers. Further analysis with greater statistical power is essential to clarify this relationship. It is important to address health equity gaps and combat systemic racism and injustice by actively investing in and procuring from minority-owned businesses.

A potential other area of analysis could also include removing/isolating the amount of money spent on suppliers that are the industry standard. For example, Epic Health Systems is one of the industry leaders in electronic medical record software applications and holds medical records of 78% of patients in the United States. Its CEO and founder is businesswoman Judidth Faulker who is was called "the most powerful woman in healthcare" by Forbes in 2013. n this dataset, as an industry leader, Epic Health Systems would still be categorized as a minority supplier due to Judith Faulkner’s ownership status. This presents an interesting dynamic, as Epic’s influence and widespread use could skew the analysis of diversity spend, highlighting funds allocated to suppliers who are industry giants. Examining the data specifically for minority suppliers who are not major players in their industries could provide valuable insights into how funds are distributed among smaller, potentially emerging minority-owned businesses and the unique challenges they face in achieving industry traction and growth.
